State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-07 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
Frequency analysis plots are added to the paper
updated required references and some paper edits to intro

\section{Introduction} This paper presents a theoretical framework for generating macroscopic neural field  mathematical models from patient specific data. The ability to estimate such models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

Parameter estimation may have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from abnormal excitatory or inhibitory mechanisms~\cite{marten2009derivation}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies. \omg{I extended this bit I am not sure how to connect it with the rest. Maybe the next paragraph can go.} 
% 
% The human brain is composed of approximately 100 billion neurons and 60 trillion synapses operating in concert and, despite decades of research, our understanding of neural dynamics remains modest. Hence research has focused on different scales of dynamics: from individual proteins, synapses, and neurons; to neuronal networks and ensembles of neuronal networks.

There has been much interest in generating physiologically plausible neural field models to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely a theoretical. Understanding the brain at this level is extremely important since this is the scale where pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested. Mathematical neural field models provide insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see \cite{Deco2008,David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999,RENNIE2000}, epileptic seizure generation \cite{DaSilva2003,Suffczynski2004,Wendling2005} and insights into other pathologies \cite{Moran2008,Schiff2009} that would be difficult to gain from experimental data alone.

@@

An estimation framework for neural field models known as dynamical causal modelling (DCM) \cite{David2003,David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range connectivity structure between the specific isolated brain regions that best explains a given data set under the Jansen and Ritt equations. Another recent publication describing a parameter estimation method with a neural field model used an unscented Kalman filter with the WC neural field equations \cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating it is possible to perform state estimation of modified WC equations, marking a first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

This paper extends the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for higher dimensional systems utilising a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been attempted. Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development in that work was to represent the spatiotemporal system as a standard state-space model with an order independent of the number of observations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of the technique to neural fields.

@@

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a reduced finite dimensional neural field model is derived in Section~\ref{ReducedModelSection}. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} then establishes conditions using spatial frequency analysis for both sensor and basis function spacing and geometry, such that the dominant dynamics of the neural field can be represented by the reduced model. Following this, the state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments. Finally, we provide concluding remarks in Section~\ref{ConclusionSection}.

\section{Neural Field Model}\label{NeuralModelSection} In this section, we describe a variant of the WC neural field model \cite{Wilson1973} that was used to create our intracortical connectivity estimator. Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. 
\begin{todo}
	{Really need to add some citations here.} 
\end{todo}
It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. 
\begin{todo}
	{Same here.} 
\end{todo}

@@

\begin{figure*}[!th] 
\subfigure[][]{\includegraphics[width=2in]{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimationError90.pdf}}\\
\caption{One snapshot of the spatial field; a: True field; b: Estimated field; c: The absolute error.}
\label{fig:FieldEstimate}
\end{figure*}
\section{Discussion}\label{DiscussionSection}
\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?} 
	\item the width of the connectivity basis functions \wtf{do you mean width of the kernel? I'm not sure about this - as the parameters increase the kernel width increases right? So maybe you mean the basis functions, but don't we estimate these widths from the frequency analysis?} here is known exactly. In practice we don't 
	\item the number of basis functions could be set higher to get a better representation of the field, however this is limited by computation complexity. \wtf{Unless you are going to specify the computational complexity, in terms of the number of basis functions, it might be better to say something like ``there is a relationship between the number of basis functions that represent the field and the computational complexity of the estimation algorithm - the more basis functions the longer it takes to estimate''} 
	\item Discuss the novelty of using the reduced model (BrainIDE) for creating patient specific models. \omg{hell yes. This should be front and centre!} 
	\item The size of the dentritic arbour is consistent in the cortex. \omg{?} 
	\item Spatial frequency of field and the relationship to the required density of electrodes/sensors required for estimation or to capture the required information without aliasing. 
	\item The flexibility of the approach 
	\item Combining this method with other to incorporate longer range hetrogeneous connectivity. \omg{awesome} 
	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
	\item Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.
	\item Estimating functional cortical connectivity via EEG promises to shed light on the hierarchy of brain regions that are involved in specific tasks, motivating the use of techniques like DCM. Other comparable methods are based on auto-regressive (AR) modelling of the EEG, MEG, or fMRI times series \todo{[refs!]}. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients using techniques like Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999}. Again, like the DCM approach, these methods estimate long range connectivity patterns that does not provide a clear relationship between the continuum field models and data.
	\item Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not estimated using DTI
\end{itemize}

>>>>

\section{Introduction} This paper presents a theoretical framework for generating macroscopic neural field mathematical models from patient specific data. There has been much interest in generating physiologically plausible neural field models to study brain dynamics at the meso/macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's meso and macro-scale dynamics remains largely a theoretical. Understanding the brain at this level is extremely important since this is the scale where pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested. Mathematical neural field models provide insights into the underlying physics and dynamics of electroencephalography (EEG) and magnetoencephalography (MEG) (see \cite{Deco2008,David2003} for recent reviews). These models have demonstrated possible mechanisms for the genesis of neural rhythms (such as the alpha and gamma rhythms) \cite{Liley1999,RENNIE2000}, epileptic seizure generation \cite{DaSilva2003,Suffczynski2004,Wendling2005} and insights into other pathologies \cite{Moran2008,Schiff2009} that would be difficult to gain from experimental data alone.

@@

An estimation framework for neural field models known as dynamical causal modelling (DCM) \cite{David2003,David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range connectivity structure between the specific isolated brain regions that best explains a given data set under the Jansen and Ritt equations. Another recent publication describing a parameter estimation method with a neural field model used an unscented Kalman filter with the WC neural field equations \cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating it is possible to perform state estimation of modified WC equations. This marks the first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

This paper extends the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for higher order systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic time constant. Until now, estimation of local intracortical connectivity structure has not been attempted. Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development in that work was to represent the spatiotemporal system as a standard state-space model with an order independent of the number of observations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of the technique to neural fields.

@@

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a reduced finite dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} establishes conditions using spatial frequency analysis for both sensor and basis function spacing and geometry, such that the dominant dynamics of the neural field can be represented by the reduced model. Following this, the state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments. Finally, we provide concluding remarks in Section~\ref{ConclusionSection}.

\section{Neural Field Model}\label{NeuralModelSection} In this section, we describe a variant of the WC or Amari style neural field model \cite{Wilson1973,Amari1977} that was used to create our intracortical connectivity estimator. Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. 
Field models are also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}

@@

\begin{figure*}[!th] 
\subfigure[][]{\includegraphics[width=2in]{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimationError90.pdf}}\\
\caption{One snapshot of the spatial field; a: True field; b: Estimated field; c: The absolute error.}
\label{fig:FieldEstimate}
\end{figure*}
\section{Discussion}\label{DiscussionSection}

Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

Parameter estimation may have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from abnormal excitatory or inhibitory mechanisms~\cite{Marten2009}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies.

\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?} 
	\item the width of the connectivity basis functions \wtf{do you mean width of the kernel? I'm not sure about this - as the parameters increase the kernel width increases right? So maybe you mean the basis functions, but don't we estimate these widths from the frequency analysis?} here is known exactly. In practice we don't 
	\item the number of basis functions could be set higher to get a better representation of the field, however this is limited by computation complexity. \wtf{Unless you are going to specify the computational complexity, in terms of the number of basis functions, it might be better to say something like ``there is a relationship between the number of basis functions that represent the field and the computational complexity of the estimation algorithm - the more basis functions the longer it takes to estimate''} 
	\item Discuss the novelty of using the reduced model (BrainIDE) for creating patient specific models. \omg{hell yes. This should be front and centre!} 
	\item The size of the dentritic arbour is consistent in the cortex. \omg{?} 
	\item Spatial frequency of field and the relationship to the required density of electrodes/sensors required for estimation or to capture the required information without aliasing. 
	\item The flexibility of the approach 
	\item Combining this method with other to incorporate longer range hetrogeneous connectivity. \omg{awesome} 
	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
	\item Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.
	\item Estimating functional cortical connectivity via EEG promises to shed light on the hierarchy of brain regions that are involved in specific tasks, motivating the use of techniques like DCM. Other comparable methods are based on auto-regressive (AR) modelling of the EEG, MEG, or fMRI times series \todo{[refs!]}. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients using techniques like Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999}. Again, like the DCM approach, these methods estimate long range connectivity patterns that does not provide a clear relationship between the continuum field models and data.
	\item Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not estimated using DTI
\end{itemize}
